Sampbias, a method to evaluate geographic sampling bias in species distribution data

### 2019-12-16 manuscript.rmd
more updates2

---
title: "Sampbias, a method to evaluate geographic sampling bias in species distribution data"
author: "Alexander Zizka^1,2,3^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4^"
bibliography: sampbias.bib
output: 
  bookdown::pdf_document2:
  # word_document:
    toc: false
    toc_depth: 1
    fig_caption: true
    number_sections: false
header-includes: |
    \usepackage{caption}
    \usepackage{fancyhdr}
    \pagestyle{fancy}
    \fancyhead[L]{sampbias accessibility bias}
    \fancyhead[R]{ } 
    \usepackage{float}
    \floatplacement{figure}{H}
    \usepackage{lineno}
    \linenumbers
    \usepackage[url=false]{biblatex}
linestretch: 2
fontsize: 12pt
always_allow_html: yes
---

@@

# Introduction
Publicly available datasets of geo-referenced species occurrences, such as provided by the Global Biodiversity Information Facility (www.gbif.org) have become a fundamental resource  in biological sciences, especially in biogeography, conservation, macroecology, and systematics. However, because these datasets are "presence-only" data, they rarely include information on collection effort. Instead they are typically not collected systematically and often compiled from a variety of sources (e.g. scientific expeditions, census counts, genetic barcoding studies, and citizen-science observations),  thus becoming subject to collection biases [@Meyer2016]. 
That is the number of data points available is biased by factors other than species' presence or abundance, including the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e. easily accessible *vs.* remote areas), and specific temporal periods ["temporal bias", i.e. wet season *vs.* dry season, @Isaac2015; @Boakes2010]. While these biases are broadly recognized, and approaches exist to account for them in some analyses (for instance for species-richness estimation [@Engemann2015] species distribution modelling [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014], occupancy modelling [@Kery2016], or abundance estimations [@Shimadzu2015]), few attempts have been made to discern among different sources of bias or to compare the strength of bias among datasets [but see @Ruete2015].

@@

# Running sampbias
A default *sampbias* analysis can be run with few lines of code in R. The main function `SamplingBias` creates an object of the class `"sampbias"`, for which the package provides a plotting and summary method. Based on a `data.frame` including species identity and geographic coordinates, sampbias provides a bias effect estimate for each gazetteer and an average bias. Additionally some options exist to provide custom gazetteers, custom distances for the bias estimation, a custom grain size of the analysis, as well as some operators for the calculation of the bias distances. A tutorial on how to use sampbias is available with the package and in the electronic supplement of this publication (Appendix S1).

```{r, eval = FALSE, tidy=TRUE}
library(sampbias)

@@

#running sampbias
example.out <- SamplingBias(x = example.in, res = 0.1)

@@

@@

# Data accessibility
The software presented here is available under a GPL-3 license. The *sampbias* R package and the source code for the shiny app are available via https://github.com/azizka/sampbias, the shiny app can be accessed at https://azizka.shinyapps.io/sampbias/. The R package includes an example dataset as well as vignettes detailing the use of the R package, the use of the shiny app and possibly warnings produced by the package (Appendix S3).

@@

# Figures

<!-- Figure 1 A sketch of the different distributions -->
```{r distdi, out.width = "7in", echo = FALSE, fig.cap = "Schematic representation of the \\textit{sampbias} algorithm. The upper panel (A-C) shows a case without accessibility bias: the observed distribution of distances from the geographic feature of interest matches the expected distribution and hence the bias function is flat. The lower panel (D-F) shows a case with bias: the expected and observed distribution differ and the bias is best represented by an exponential function.", eval = TRUE}
knitr::include_graphics(path = "ms_figures/Distance_distribution_example.png")
```

<!-- Figure 2 shiny interface -->
```{r GUI, out.width = "7in", echo = FALSE, fig.cap = "The interface of the sampbias shiny app. With a navigation panel on the left and visualization of the results on the right.", eval = TRUE}
knitr::include_graphics(path = "ms_figures/bild_gui_sampbias.png")
```

<!-- Figure 3 empirical results -->
```{r empirical, out.width = "7in", echo = FALSE, fig.cap = "The spatial projection of the accessibility bias in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo from www.gbif.org. \\textit{Sampbias} finds a strong biasing effect by roads, cities and lower by airports, but, surprisingly, no effect of rivers.", eval = TRUE}
knitr::include_graphics(path = "ms_figures/visualization_rpackage_application_githubtutorial.jpg")
```

@@

Appendix S2 - Tutorial running the sampbias shiny app

Appendix S3 - Possible warnings and their solutions

>>>>

---
title: "Sampbias, a method to evaluate geographic sampling bias in species distribution data"
author: "Alexander Zizka^1,2,3^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4^"
bibliography: sampbias.bib
output: 
  bookdown::pdf_document2:
  # word_document:
    toc: false
    toc_depth: 1
    fig_caption: true
    number_sections: false
header-includes: |
    \usepackage{caption}
    \usepackage{fancyhdr}
    \pagestyle{fancy}
    \fancyhead[L]{sampbias accessibility bias}
    \fancyhead[R]{\thepage}
    \fancyfoot{}
    \usepackage{float}
    \floatplacement{figure}{H}
    \usepackage{lineno}
    \linenumbers
    \usepackage[url=false]{biblatex}
linestretch: 2
fontsize: 12pt
always_allow_html: yes
---

@@

# Introduction
Publicly available datasets of geo-referenced species occurrences, such as provided by the Global Biodiversity Information Facility (www.gbif.org) have become a fundamental resource  in biological sciences, especially in biogeography, conservation, macroecology, and systematics. However, because these datasets are "presence-only" data, they rarely include information on collection effort. Instead they are typically not collected systematically and often compiled from a variety of sources (e.g. scientific expeditions, census counts, genetic barcoding studies, and citizen-science observations),  thus becoming subject to collection biases [@Meyer2016].

That is the number of data points available is biased by factors other than species' presence or abundance, including the under-sampling of specific taxa ("taxonomic bias", e.g., birds *vs.* nematodes), specific geographic regions ("geographic bias", i.e. easily accessible *vs.* remote areas), and specific temporal periods ["temporal bias", i.e. wet season *vs.* dry season, @Isaac2015; @Boakes2010]. While these biases are broadly recognized, and approaches exist to account for them in some analyses (for instance for species-richness estimation [@Engemann2015] species distribution modelling [@Stolar2015;@Beck2014; @Fithian2014; @Warren2014; @Boria2014; @Varela2014; @Fourcade2014], occupancy modelling [@Kery2016], or abundance estimations [@Shimadzu2015]), few attempts have been made to discern among different sources of bias or to compare the strength of bias among datasets [but see @Ruete2015].

@@

# Running sampbias
A default *sampbias* analysis can be run with few lines of code in R. The main function `calculate_bias` creates an object of the class `"sampbias"`, for which the package provides a plotting and summary method. Based on a `data.frame` including species identity and geographic coordinates, sampbias provides a bias effect estimate for each gazetteer and an average bias. Additionally some options exist to provide custom gazetteers, custom distances for the bias estimation, a custom grain size of the analysis, as well as some operators for the calculation of the bias distances. A tutorial on how to use sampbias is available with the package and in the electronic supplement of this publication (Appendix S1).

```{r, eval = FALSE}
library(sampbias)

@@

#running sampbias
example.out <- calculate_bias(x = example.in, res = 0.1)

@@

#project in space
proj <- project_bias(example.out)
map_bias(proj)

@@

# Data accessibility
The software presented here is available under a GPL-3 license. The *sampbias* R package and the source code for the shiny app are available via https://github.com/azizka/sampbias. The R package includes an example dataset as well as vignettes detailing the use of the R package, the use of the shiny app and possibly warnings produced by the package (Appendix S2).

@@

# Figures
<!-- mpirical results -->
```{r empirical, out.width = "6in", echo = FALSE, fig.cap = "The spatial projection of the accessibility bias in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo from www.gbif.org. A) bias weights, B) projection of the expected number of occurrences given the sampbias model. \\textit{Sampbias} finds the strongest biasing effect for cities.", eval = TRUE}
knitr::include_graphics(path = "ms_figures/figure_empirical_results.jpg")
```

@@

Appendix S2 - Possible warnings and their solutions
